---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Providing data access via RESTful APIs
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
library(tidyverse)

opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```


## Time table Tuesday, February 18

``` {r ttableTue, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-02-18")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>% 
  row_spec(3, background = trr266_lightpetrol)
```


## What are RESTful APIs?

- In short: The build blocks of the World-Wide Web
- Somewhat longer: Application Programming Interface(s) that allow clients to
access services from servers in a stateless and cache-able way using uniform
interfaces
- This allows clients to be developed independently from servers and data to be 
shared efficiently across dynamic networks


## Sharing your data via a RESTful HTTP API: Meet Plumber

\begin{center}
\includegraphics[width=0.5\textwidth]{media/plumber-broken.png} 
\end{center}