---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Assignments and wrap-up 
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)

opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```


## Time table October 11

``` {r ttable, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-10-11")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>%
  row_spec(7, background = trr266_lightpetrol)
```


## Assignments

- If you want to receive credits (6 ECTS) for this course you need to
  hand in three individual and one group assignment on time.
  
- Individual assignment: Coding sprints (three week period, scattered throughout 
  the term, 5 in total)
  
- Group projects: One larger real-life task. The project needs to generate an
  asset for the open science community.
  
- Groups should consist of not more than five people. Please try to
  self-coordinate groups based on skills and domain interest. Let me know when 
  you need help. Groups should be formed by the end of next week, preferably 
  today.

- While the coding sprints will be determined and introduced by me, the group
  assignments will be negotiated with each group. I have some ideas but am 
  open for suggestions.

  

## Your first sprint assignment

\footnotesize

- Provide code that produces a GIF documenting something interesting based on 
  World Bank data. Communicating the increasing income gap between Africa and the 
  rest of the world is an option but you can also choose a different topic.

- An example:
  https://twitter.com/DLeonhardt/status/1181004566088814594

- Data: `worldbank` dataset from `ExPanDaR` package or 
  https://data.worldbank.org
  
- Your solution has to contain code and documentation so that it can
  be reproduced in our development environment. Data should _not_ be included
  but the code should pull it from publicly available resources.
  
- Deadline for submission: Monday, Oct 28, 9am. Please only commit locally prior
  to the deadline (maybe in a separate branch?) and submit by email.

- After the deadline, feel free to push your code to your remote
  fork for everybody to enjoy.
  
\normalsize

## Our next physical meeting: Feb 17/18, 2020

Coming up: Even more exciting topics

- Group presentations of their work
- Testing
- Profiling
- Data scraping
- Interactive visualization
- Simulations and effect size analysis
- Explore your researcher degrees of freedom
- Providing data access via RDBMS and APIs

\vspace{12pt}
\center

Mark your calendars and make sure that you receive GitHub notifications by "watching" the `sposm` repository!

# Thank you very much!

