---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Functional versus object-oriented programming
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
library(ggmap)
opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```

## Time table October 10

``` {r ttable, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-10-10")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>%
  row_spec(8, background = trr266_lightpetrol)
```


## Functional programming versus scripting

* Many statistical programming languages (EViews, SAS, Stata, R to some 
extent) are in essence scripting languages.
* Scripts are closely connected to imperative programming ("Shut up and do what 
I tell you!")
* Scripts are hard to read, tend to become inefficient, and are hard to reuse
* "If you copy + paste your (own) code a lot, you are a bad programmer"


## The key idea of functional programming

* Functional programming is declarative in nature: Your functions describe
what to do. The implementation is hidden from the user.
* A function takes arguments, processes them and returns results
* A _pure function_ is a function where
  - the result of the function depends only on its arguments and
  - that generates no _side effects_
* Pure functions are *referentially transparent*, meaning that they can be 
replaced with their return value without changing the program
* In real-life coding, many functions are not referentially transparent. This
makes writing code easier and reading code harder 


## Functions in R

* Functions have three components: 
  - `formals()`: The arguments that you call the function with 
  - `body()`: The code that the function executes
  - `environment()`: The place where the function can look for objects
* Functions are objects, just like about anything else in R
* Internally, they are called `closures`. Knowing this can be helpful to 
decipher error messages!


## Chaining functions in R: Intermediate objects

Readable but tedious
\vspace{36pt}
\footnotesize
```{r, eval = FALSE}
df <- read_csv("data/sub.csv")
df <- select(df, cik, name)
df <- distinct(df)
count_sec_reg <- nrow(df)
sprintf("There are %d registrants", count_sec_reg)
```
\normalsize

## Chaining functions in R: Nesting

Concise but a pain in the eye
\vspace{36pt}
\footnotesize
```{r, eval = FALSE}
sprintf(
  "There are %d registrants",
  nrow(distinct(select(read_csv("data/sub.csv"), cik, name)))
)
```
\normalsize

## Chaining functions in R: Piping

The tidy way (read `%>%` as "and then") but harder to debug
\vspace{36pt}
\footnotesize
```{r, eval = FALSE}
read_csv("data/sub.csv") %>%
  select(cik, name) %>%
  distinct() %>%
  nrow() -> count_sec_reg

sprintf("There are %d registrants", count_sec_reg)
```
\normalsize


## Scoping I

What does this code snippet return?
\vspace{36pt}
\footnotesize
```{r, eval = FALSE}
x <- 10

my_func <- function() {
  x <- 20
  x
}

c(my_func(), x)
```
\normalsize


## Scoping I

What does this code snippet return?
\vspace{36pt}
\footnotesize
```{r}
x <- 10

my_func <- function() {
  x <- 20
  x
}

c(my_func(), x)
```
\normalsize


## Scoping II

What does this code snippet return?
\vspace{36pt}
\footnotesize
```{r, eval = FALSE}
x <- 10
y <- 5

my_func <- function() {
  x <- 20
  x*y
}

my_func()
```
\normalsize


## Scoping II

What does this code snippet return?
\vspace{36pt}
\footnotesize
```{r}
x <- 10
y <- 5

my_func <- function() {
  x <- 20
  x*y
}

my_func()
```
\normalsize


## Scoping III

What does this code snippet return?
```{r include = FALSE}
rm(x, y)
```
\vspace{36pt}
\footnotesize
```{r, eval = FALSE}
my_second_func <- function(x) {
  y <- x
}

my_func <- function(x) {
  x*y
}

my_second_func(5)
my_func(10)
```
\normalsize


## Scoping III

What does this code snippet return?
\vspace{36pt}
\footnotesize
```{r, error = TRUE}
my_second_func <- function(x) {
  y <- x
}

my_func <- function(x) {
  x*y
}

my_second_func(5)
my_func(10)
```
\normalsize


## Scoping IV

What does this code snippet return?
\vspace{36pt}
\footnotesize
```{r, eval = FALSE}
my_second_func <- function(x) {
  y <<- x
}

my_func <- function(x) {
  x*y
}

my_second_func(5)
my_func(10)
```
\normalsize


## Scoping IV

What does this code snippet return?
\vspace{36pt}
\footnotesize
```{r,}
my_second_func <- function(x) {
  y <<- x
}

my_func <- function(x) {
  x*y
}

my_second_func(5)
my_func(10)
```
\normalsize


## For extra credit

\center
Which of these functions are pure, which are not? Why? 

## Recursions: Functions can be very helpful

``` {r tower-of-hanoi, echo = FALSE, fig.width = 4, fig.height = 1.5, out.width = "0.5\\textwidth"}
ggplot() +
  geom_leg(aes(x = 0, y = 0, xend = 10, yend = 0), size = 2.5) +
  geom_leg(aes(x = 2, y = 0, xend = 2, yend = 0.5), size = 2.5) +
  geom_leg(aes(x = 5, y = 0, xend = 5, yend = 0.5), size = 2.5) +
  geom_leg(aes(x = 8, y = 0, xend = 8, yend = 0.5), size = 2.5) +
  geom_rect(aes(xmin = 0.5, xmax = 3.5, ymin = 0.02, ymax = 0.07), 
            color = NA, fill = trr266_petrol) +
  geom_rect(aes(xmin = 1, xmax = 3, ymin = 0.08, ymax = 0.13), 
            color = NA, fill = trr266_red) +
  geom_rect(aes(xmin = 1.5, xmax = 2.5, ymin = 0.14, ymax = 0.19), 
            color = NA, fill = trr266_yellow) + theme_void()
```
\vspace{12pt}
\footnotesize
```{r, eval = FALSE}
tower <- function(n, from_peg, to_peg, aux_peg) {
  if(n == 0) return(invisible())
  tower(n - 1, from_peg, aux_peg, to_peg)
  message(sprintf("Moving piece %d from %s to %s ...", 
                  n, from_peg, to_peg), appendLF = FALSE)
  tower(n - 1, aux_peg, to_peg, from_peg)
}
tower(3, 'F', 'T', 'A')
```
\normalsize


## Recursions: Functions can be very helpful

\scriptsize
```{r, message = TRUE}
tower <- function(n, from_peg, to_peg, aux_peg) {
  if(n == 0) return(invisible())
  tower(n - 1, from_peg, aux_peg, to_peg)
  message(sprintf("Moving piece %d from %s to %s ...", 
                  n, from_peg, to_peg))
  tower(n - 1, aux_peg, to_peg, from_peg)
}
tower(3, 'F', 'T', 'A')
```
\normalsize
See https://www.youtube.com/watch?v=YstLjLCGmgg for animation


## Object oriented programming

* Much more common in Python that in R, object oriented programming
encapsulates data and functions (aka as _methods_ in the OOP world) in _classes_
* Methods can be _overloaded_ by _inheriting_ classes
* Tends to make code more consistent and easier to maintain/extend
* Makes it easier for code to modify data (makes objects more _mutable_), 
something that people in statistical programming are generally not very fond off
* How does this look like: Let's have a quick look at a last toy example
  - `code/show_fs_oop.py` versus
  - `code/show_fs_fp.R`
