---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Writing readable and reusable code
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
library(ggmap)
opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```

## Time table October 11

``` {r ttable, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-10-11")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>%
  row_spec(1, background = trr266_lightpetrol)
```


## Most of us have been going through loads of this

\tiny
```
clear all
import excel "clean_data\interviews_quant_data_final_15-09-16.xlsx", sheet("MainData") firstrow
gen treated_case = .
replace treated_case = 1 if treatment == "1" & (obj_val == 1 & obj_perf_mgt == 0 & incent_finacc == 0 & incent_intprod == 1)
replace treated_case = 2 if treatment == "2" & (obj_val == 0 & obj_perf_mgt == 1 & incent_finacc == 0 & incent_intprod == 1)
replace treated_case = 3 if treatment == "3" & (obj_val == 1 & obj_perf_mgt == 0 & incent_finacc == 1 & incent_intprod == 0)
replace treated_case = 4 if treatment == "4" & (obj_val == 0 & obj_perf_mgt == 1 & incent_finacc == 1 & incent_intprod == 0)
gen treated = treated_case != .
gen exp_pilot = (treatment == "C" | treatment == "V")
gen failed_man_checks = !exp_pilot & treated_case == .
gen highest_college_ed = "none"
replace highest_college_ed = "bachelor" if bachelor==1
replace highest_college_ed = "masters" if masters==1
replace highest_college_ed = "phd" if phd==1
gen acctg_expertise = "none"
replace acctg_expertise = "intermediate (bus/econ and/or CFA)" if cfa | business_econ
replace acctg_expertise = "advanced (actg degree or acctg responsibility in current position)" if acctg_qual | fin_rep_job_resp
gen occupation = ""
replace occupation = "fund manager" if (fund_manager | (value_funds_direct_mgmtM >0 & value_funds_direct_mgmtM < .))
replace occupation = "analyst buy-side" if (occupation != "fund manager" & analyst  == 1 & buy_side ==1)
replace occupation = "analyst sell-side" if (occupation == "" & analyst  == 1 & sell_side ==1)
replace occupation = "other" if occupation == ""
gen asset_focus = ""
replace asset_focus = "equity only" if (equity == 1 & debt == 0)
replace asset_focus = "debt only" if (equity == 0 & debt == 1)
replace asset_focus = "equity and debt" if (equity == 1 & debt == 1)
replace asset_focus = "other" if (equity == 0 & debt == 0)
gen firm_focus = ""
replace firm_focus = "public only" if (public == 1 & private == 0)
replace firm_focus = "private only" if (public == 0 & private == 1)
replace firm_focus = "public and private" if (public == 1 & private == 1)
replace firm_focus = "other" if (public == 0 & private == 0)
```
\normalsize


## Open Science needs readable code!

* To make an impact, code needs to be digestible for the reader
* Besides a well-designed project setup, accessible code is a key 
ingredient to enable others to contribute to your work

\vspace{48pt}

\center
But what are the key ingredients to make your code readable and reusable?

## Rule #1: Use a style guide

Each serious programming language has established style guides, e.g.:

\vspace{24pt}

- https://google.github.io/styleguide/ for various languages 
- https://style.tidyverse.org for R
- https://www.python.org/dev/peps/pep-0008/ The "official" python style guide

\vspace{24pt}
\centering
Linters can help with getting your code in shape


## Rule #2: Write code as you speak

\footnotesize
```{r, eval = FALSE}
# Bad:

df <- read_csv("data/sub.csv")
x <- nrow(df[!duplicated(df[, 3:4]), 3:4])
sprintf("There are %d registrants", x)

# Good: 

read_csv("data/sub.csv") %>%
  select(cik, name) %>%
  distinct() %>%
  nrow() -> count_sec_reg

sprintf("There are %d registrants", count_sec_reg)
```
\normalsize


## Rule #3: Use functions for reusable steps

\footnotesize
```{r, eval = FALSE}
count_distinct_obs <- function(df, ...) {
  as_tibble(df) %>%
    select(...) %>%
    distinct() %>%
    nrow()
}

read_csv("data/sub.csv") %>%
  count_distinct_obs(cik, name) -> count_sec_reg

sprintf("There are %d registrants", count_sec_reg)
```
\normalsize


## Rule #4: Keep functions short and indentation levels low

\scriptsize
```{r, eval = FALSE}
# Bad:

if (use_server_ok()) {
  if (ping_server_ok()) {
    if (connect_server_ok()) {
      df <- read_server_data()
    } else {
      if (fall_back_ok()) {
        df <- read_fall_back()
      } else stop("No data")
    }
  } else {
    if (fall_back_ok()) {
      df <- read_fall_back()
    } else stop("No data")
  }
} else {
  if (fall_back_ok()) {
    df <- read_fall_back()
  } else stop("No data")
}
do_something_with_data(df)
```
\normalsize


## Rule #4: Keep functions short and indentation levels low

\scriptsize
```{r, eval = FALSE}
# Good:

test_server_ok <- function{
  if (!ping_server_ok()) return(FALSE)
  if (!connect_server_ok()) return(FALSE)
  return(TRUE)
}

get_data <- function() {
  if (use_server_ok()) {
    if (test_server_ok()) return(read_server_data())
  }
  if (fall_back_ok()) return(read_fall_back())
  else stop("No data")
} 

get_data() %>%
  do_something_with_data()
```
\normalsize


## Rule #5: Check for errors

\footnotesize
```{r, collapse = TRUE}
count_distinct_obs <- function(df, ...) {
  as_tibble(df) %>%
    select(...) %>%
    distinct() %>%
    nrow()
}

count_distinct_obs(42)
```
\normalsize


## Rule #5: Check for errors

\scriptsize
```{r, error = TRUE, collapse = TRUE}
count_distinct_obs <- function(df, ...) {
  if (!is.data.frame(df)) {
    stop("First parameter must be a data frame", call. = FALSE) 
  }
  if (missing(...)) {
    stop("Must have at least one column variable", call. = FALSE)
  }

  df %>%
      select(...) %>%
      distinct() %>%
      nrow()
}

count_distinct_obs(42)
count_distinct_obs(data.frame(a = 1:10))
count_distinct_obs(data.frame(a = 1:10), b)
count_distinct_obs(data.frame(a = 1:10), a)
```
\normalsize


## Rule #6: Don't document code, document functions

\tiny
```{r, eval = FALSE}
#' @title Counts the destinct observations for a sub-set of a data frame
#'
#' @description
#' Reads a data frame and counts the number of distinct observations based 
#' on a subset of provided columns.
#' @param df Data frame
#' @param ... The list of columns to include (cannot be empty)
#' @return The number of observations that are not identical across the selected
#' columns
#'
#' @examples
#' df <- data.frame(a = rnorm(5), b = 1, c = c(1, 2, 1, 2, 3))
#' count_distinct_obs(df, b, c)
#' @export

count_distinct_obs <- function(df, ...) {
  if (!is.data.frame(df)) {
    stop("First parameter must be a data frame", call. = FALSE) 
  }
  if (missing(...)) {
    stop("Must have at least one column variable", call. = FALSE)
  }

  df %>%
      select(...) %>%
      distinct() %>%
      nrow()
}
```
\normalsize
