---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Explore your researcher degrees of freedom
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{adjustbox}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
- \definecolor{trr266petrol}{RGB}{27,138,143}
- \definecolor{trr266blue}{RGB}{110,202,226}
- \definecolor{trr266yellow}{RGB}{255,180,59}
- \definecolor{trr266red}{RGB}{148,70,100}
- \newcommand{\code}{\texttt}
- \newcommand{\bcenter}{\begin{center}}
- \newcommand{\ecenter}{\end{center}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
library(tidyverse)
library(ExPanDaR)
library(rdfanalysis)

opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```

```{r prepare_data, include=FALSE}
read_csv("../raw_data/restrans_data.csv", col_types = cols()) %>%
  mutate_at(c("country", "year"), as.factor) %>%
  mutate_at(c("life_expectancy", "gdp_capita",
              "mn_yrs_school", "unemployment"), list(ln = log)) -> smp_all

smp_no_na <- smp_all %>% na.omit

var_def <- read_csv("../raw_data/restrans_data_def.csv", col_types = cols())

for (var in c("life_expectancy", "mn_yrs_school", 
              "mn_yrs_school", "unemployment")) {
  var_def <- rbind(var_def,
                   list(paste0(var, "_ln"), 
                        paste("Log-transformed version of", var), "numeric", 1))
}
```

## Time table Tuesday, February 18

``` {r ttableTue, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-02-18")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>% 
  row_spec(1, background = trr266_lightpetrol)
```


## Motivation

\begin{center}
\includegraphics[height=0.7\textheight]{media/leamer_aer_1983_37f.png} \\
Leamer (AER 1983: 37f.)
\end{center}


## Slightly more systematic ...

\begin{center}
\includegraphics[height=0.8\textheight]{media/wicherts_et_al_2016_table_1_p3.png} \\
Wicherts et al (Frontiers in Psychology 2016: 3)
\end{center}


## Objective

Development of a statistical computing framework for a "multiverse analysis" (Steeger et al., Persp on Psych Sci 2016) that

- supports research design development separate from data,
- supports pre-registering of observational data research designs,
- facilitates a priori power analysis,
- promotes unit testing,
- generates well documented and easily portable code,
- makes researcher degrees of freedom (Simmons et al., Psyc Science 2011) explicit in code, and
- allows for rigorous robustness checking by exhausting all these degrees of freedom algorithmically. 

Note: Code based walk-through on what follows: https://joachim-gassen.github.io/rdfanalysis/articles/analyzing_rdf.html

## Workflow

1. Sketch the \textcolor{black}{\code{design}}, the \textcolor{trr266yellow}{\code{result}} and the required \textcolor{trr266petrol}{\code{steps}} to generate it
2. Develop test code for \textcolor{gray}{\code{input}} data
3. Develop function to simulate \textcolor{gray}{\code{input}} data. Verify it with test code
4. For each \textcolor{trr266petrol}{\code{step}}
    - Draft \textcolor{trr266petrol}{\code{step\_description}} and develop test code
    - Identify and implement potential \textcolor{trr266blue}{\code{choices}}. Document them by providing the \textcolor{trr266blue}{\code{choice\_description}}
    - Verify that \textcolor{trr266petrol}{\code{step}} passes tests
5. Verify that \textcolor{black}{\code{design}} produces unbiased \textcolor{trr266yellow}{\code{results}} consistent with simulated effect sizes
6. Run power simulation with your a priori estimation of effect size
7. Register your priors about choices by assigning \textcolor{gray}{\code{weights}}
8. Run \textcolor{black}{\code{design}} with real data and interpret \textcolor{trr266yellow}{\code{result}}


## Implementation of a \textcolor{black}{\code{design}}

* A `design` consists of  \textcolor{trr266petrol}{\code{steps}}. 
* Each \textcolor{trr266petrol}{\code{step}} reads the \textcolor{gray}{\code{output}} of the prior \textcolor{trr266petrol}{\code{step}} as \textcolor{gray}{\code{input}} and can contain any number of 
    - \textcolor{trr266blue}{\code{discrete choices}} and
    - \textcolor{trr266red}{\code{continuous choices}}
* The output of the final \textcolor{trr266petrol}{\code{step}} provides the \textcolor{trr266yellow}{\code{result}}
* By referring to the \textcolor{gray}{\code{protocol}} of prior \textcolor{gray}{\code{choices}}, this set-up can be used to design arbitrarily complex forking paths (Gelman and Loken, Am Scientist 2011)


## A case study: Preston (1975)

\begin{center}
\includegraphics[height=0.7\textheight]{media/preston_1975_p235.png}
Preston (1975): The Changing Relation between Mortality and level
of Economic Development, Population Studies (29): 235.
\end{center}


## Descriptive Statistics

```{r tab_descriptives_sample, echo = FALSE, results="asis", out.height="\\textheight"}
df <- smp_no_na %>% 
  select(life_expectancy, gdp_capita, mn_yrs_school, unemployment) %>%
  mutate(gdp_capita = gdp_capita/1000)

t <- prepare_descriptive_table(df)

desc_rnames <- c(
  "\\textit{Life expectancy}", 
  "\\textit{GDP per capita}", 
  "\\textit{Years of schooling}", 
  "\\textit{Unemplyoment rate}"
)
rownames(t$df) <- desc_rnames
names(t$df)[c(5,7)] <- c("25 \\%", "75 \\%")
kable(t$df, digits = c(0, 1, 1, 1, 1, 1, 1, 1), format = "latex",
      format.args = list(decimal.mark = ".", big.mark = ",", 
                         scientific=FALSE),
      booktabs = TRUE, escape = FALSE, linesep = "") -> kab_latex

lat_tab <- unlist(strsplit(kab_latex, "\n"))
lat_tab[3] <- "\\\\[-1.8ex]\\hline \\hline \\\\[-1.8ex]"
lat_tab[5] <- "\\hline\\\\[-1.8ex]"
lat_tab[length(lat_tab) - 1] <- "\\\\[-1.8ex]\\hline \\hline \\\\[-1.8ex]"

latex_tab <- c("\\begin{threeparttable}",
               lat_tab[2:length(lat_tab)],
               "\\begin{tablenotes}[flushleft]",
               "\\setlength{\\labelsep}{0pt}",
               sprintf("\\item Note: The data is obtained from the World Bank and the Wittgenstein Center. The sample covers %d countries and the period %d to %d. \\textit{GDP per captia} are in constant 2010 thousand U.S. dollars.", length(unique(smp_no_na$country)),
                       min(as.numeric(as.character(smp_no_na$year))), 
                       max(as.numeric(as.character(smp_no_na$year)))),
               "\\end{tablenotes}",
               "\\end{threeparttable}")

cat("\\centering \\adjustbox{max height=\\dimexpr\\textheight-6cm\\relax, max width=\\textwidth}{")
cat(paste(latex_tab, collapse = "\n"))  
cat("}")
```


## Regression Results

```{r estimate_models, include=FALSE}
res <-  prepare_regression_table(
  smp_no_na,
  dvs = rep("life_expectancy", 4),
  idvs = list(
    "gdp_capita_ln", 
    c("gdp_capita_ln", "mn_yrs_school_ln", "unemployment_ln"),
    c("gdp_capita_ln", "mn_yrs_school_ln", "unemployment_ln"),
    c("gdp_capita_ln", "mn_yrs_school_ln", "unemployment_ln")
  ),
  feffects = list("", "", "year", c("country", "year")),
  clusters = list("", "", "year", c("country", "year")),
  format = "latex"
)
min_est = min(confint(res$models[[1]]$model)[2, 1],
              confint(res$models[[2]]$model)[2, 1],
              confint(res$models[[3]]$model)[1, 1],
              confint(res$models[[4]]$model)[1, 1])
              
max_est = max(confint(res$models[[1]]$model)[2, 2],
              confint(res$models[[2]]$model)[2, 2],
              confint(res$models[[3]]$model)[1, 2],
              confint(res$models[[4]]$model)[1, 2])
```

```{r tab_regression_results, echo = FALSE, results='asis'}

vars_out <- c("gdp\\_capita\\_ln", "mn\\_yrs\\_school\\_ln", 
              "unemployment\\_ln", "country", "year")
vars_in <- c("ln(\\textit{GDP per capita})", "ln(\\textit{Years of schooling})",
             "ln(\\textit{Unemployment rate})", "Country", "Year")
tb <- res$table

for (i in 1:length(vars_out)) {
  tb <- str_replace_all(tb, fixed(vars_out[i]), vars_in[i])
}


tb[10] <- " & \\multicolumn{4}{c}{Dependent variable: \\textit{Life expectancy}} \\\\ "  
latex_tab <- c("\\small",
               "\\begin{threeparttable}",
               "\\setlength{\\tabcolsep}{2pt}",
               "\\begin{tabular}{lcccccccc}",
               tb[8:11], tb[13:27], tb[29:35],
               "\\end{tabular}",
               "\\begin{tablenotes}[flushleft]",
               "\\setlength{\\labelsep}{0pt}",
               "\\item Note: The dependent variable is the average life expectancy at birth in years. OLS coefficients are reported together with standard errors in parentheses. $^{*}$/$^{**}$/$^{***}$ indicate two-sided significance levels of 10/5/1 \\%, respectively.",
               "\\end{tablenotes}",
               "\\end{threeparttable}")

cat("\\centering \\adjustbox{max height=\\dimexpr\\textheight-6cm\\relax, max width=\\textwidth}{")
cat(paste(latex_tab, collapse = "\n"))  
cat("}")
```


## The steps and their discrete/continuous choices

\begin{center}
\includegraphics[height=0.85\textheight]{media/rdf_design.png}
\end{center}


## Optional: A power analysis on simulated data I

\begin{center}
\includegraphics[width=0.85\textwidth]{media/rdf_check_power_1.png} \\
Power estimation based on an assumed effect size of one year for a 10 % 
increase of GDP per capita. Each country has 15 years of data. n: Number of 
countries in sample.
\end{center}


## Optional: A power analysis on simulated data II

\begin{center}
\includegraphics[width=0.85\textwidth]{media/rdf_check_power_2.png} \\
Power estimation based on an assumed effect size of one year for a 10 % 
increase of GDP per capita. Each country has 15 years of data. n: Number of 
countries in sample.
\end{center}


## Estimate the effect size on real data

``` {r exhaust_weighted_ests, include = "FALSE", results = "hide"}
load(url("https://joachim-gassen.github.io/data/rdf_ests.RData"))
```
``` {r table_weighted_results, echo = FALSE, results='asis'}

kable(weighted_ests, digits = 3, booktabs = TRUE, linesep = "",
      format = "latex") -> kab_latex
lat_tab <- unlist(strsplit(kab_latex, "\n"))

cat("\\centering \\adjustbox{max height=\\dimexpr\\textheight-6cm\\relax, max width=\\textwidth}{")
cat(paste(lat_tab, collapse = "\n"))  
cat("}")

```

And the weighted average estimate is:

``` {r estimate_weighted_effect, echo = FALSE}
weighted_eff <- calculate_weighted_estimate(weighted_ests, "est", "lb", "ub")
kable(as.data.frame(weighted_eff), digits = 3, booktabs = TRUE, linesep = "")
```


## Some but not all of the researcher degrees of freedom

```{r SepcCurve, cache = TRUE, echo = FALSE, fig.align="center", out.width="\\textwidth"}
df <- ests %>%
  filter(na.omit == "no",
         outlier_cutoff == 0 | outlier_cutoff == 0.01 | outlier_cutoff == 0.05,
         cluster == "ctryyear") %>%
  select(-na.omit, -cluster)
attr(df, "choices") <- 1:5
plot_rdf_spec_curve(df, "est", "lb", "ub", 
                    est_label = expression(Delta*" Life expectancy (yrs)"), pt_size = 0.5,
                    choice_ind_point = FALSE) 
```


## And an interactive variant of it

\bcenter

https://jgassen.shinyapps.io/shiny_rdf_spec_curve/

\ecenter

## For further info and code

\bcenter

https://joachim-gassen.github.io/rdfanalysis

\ecenter
