---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Data wrangling and visualization fundamentals
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
#devtools::install_github("bergant/datamodelr")
library(ExPanDaR)
library(cowplot)
library(showtext)

font_add_google("PT Sans", "PTSans")
opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```

## Time table October 11

``` {r ttable, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-10-11")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>%
  row_spec(6, background = trr266_lightpetrol)
```


## Disclaimer

\center A lot of what follows --- including but not limited to the figures --- 
are borrowed from Claus O. Wilke (2019): Fundamentals of Data Visualization, https://serialmentor.com/dataviz/index.html


## Data wrangling/cleaning

* We already talked a lot about data wrangling in the last session

* Special issues like data scraping will be covered in the 2nd course
block in February

* Instead, I would like to focus on one issue that I believe to be
paramount in applied econometrics: miss-coded data


## Miss-coded data

* Data in econometrics only rarely has unambiguous measurement properties

* Money is opinion

* Magnitudes of measures are heavily affected by the size of the data generating
unit (firm, countries, etc.)

* A common remedy is to size-adjust measures by deflating but this sometimes
tend to introduce new problems


## A macro example that we all know

```{r, echo = FALSE, cache = TRUE, fig.width=4, fig.height=2, out.width="0.9\\textwidth"}
data("worldbank_var_def")
data("worldbank_data_def")
data("worldbank")

wb <- calc_variables(worldbank,
                     worldbank_var_def$var_name,
                     worldbank_var_def$var_def,
                     worldbank_var_def$type,
                     worldbank_var_def$can_be_na)


ggplot(data = wb, aes(x = gdp)) +
  geom_histogram(fill = trr266_lightpetrol) +
  labs(x = "GDP") + theme_minimal()
```

\scriptsize

Definition: `r worldbank_data_def$var_def[23]`

\normalsize


## A macro example that we all know

```{r, echo = FALSE, cache = TRUE, fig.width=4, fig.height=2, out.width="0.9\\textwidth"}
ggplot(data = wb, aes(x = log(gdp))) +
  geom_histogram(fill = trr266_lightpetrol) +
  labs(x = "ln(GDP)") + theme_minimal()
```

\scriptsize

Definition: `r worldbank_data_def$var_def[23]`

\normalsize

## The resulting deflator problem

```{r, echo = FALSE, cache = TRUE, fig.width=4, fig.height=3, out.width="0.9\\textwidth", warning=FALSE}
ggplot(data = wb, aes(x = log(gdp), y = imports_gdp)) +
  geom_point(size = 0.5, alpha = 0.5, color = trr266_lightpetrol) +
  labs(x = "ln(GDP)", y = "Imports % GDP") + theme_minimal()
```


## Flawed variable definitions

\center
$ROE_{i,t} = \frac{NI_{i,t}}{0.5(BVE_{i,t-1} + BVE_{i,t})}$

```{r, echo = FALSE, cache = TRUE, fig.width=4, fig.height=2.5, out.width="0.8\\textwidth", warning=FALSE}
data("russell_3000")
ggplot(data = russell_3000, aes(x = roe, group = equity > 0)) +
  geom_histogram(fill = trr266_lightpetrol) +
  labs(x = "Return over equity") + theme_minimal()
```

\raggedright
\scriptsize

Definition: `r russell_3000_data_def %>% filter(var_name == "roe") %>% select(var_def)`.
Data: Russel 3000 U.S. Index firms, fiscal years 2014-2016, as provided by the
ExPanDaR R package, https://joachim-gassen.github.io/ExPanDaR/.

\normalsize


## Flawed variable definitions: Winsorization

\center
$ROE_{i,t} = \frac{NI_{i,t}}{0.5(BVE_{i,t-1} + BVE_{i,t})}$

```{r, echo = FALSE, cache = TRUE, fig.width=4, fig.height=2.5, out.width="0.8\\textwidth", warning=FALSE}
russell_3000$roe <- treat_outliers(russell_3000$roe, percentile = 0.01) 
ggplot(data = russell_3000, aes(x = roe)) +
  geom_histogram(fill = trr266_lightpetrol) +
  labs(x = "Return over equity") + theme_minimal()
```

\raggedright
\scriptsize

Definition: `r russell_3000_data_def %>% filter(var_name == "roe") %>% select(var_def)`.
Data: Russel 3000 U.S. Index firms, fiscal years 2014-2016, as provided by the
ExPanDaR R package, https://joachim-gassen.github.io/ExPanDaR/.

\normalsize


## Flawed variable definitions: Unintended side-effects

\center
$ROE_{i,t} = \frac{NI_{i,t}}{0.5(BVE_{i,t-1} + BVE_{i,t})}$

```{r, echo = FALSE, cache = TRUE, fig.width=4, fig.height=2.5, out.width="0.8\\textwidth", warning=FALSE}
russell_3000$roe <- treat_outliers(russell_3000$roe, percentile = 0.01) 
ggplot(data = russell_3000, aes(x = roe, fill = equity > 0)) +
  geom_histogram() +
  labs(x = "Return over equity") + 
  scale_fill_manual(values = c(trr266_yellow, trr266_lightpetrol)) +
  theme_minimal() + theme(legend.position = c(0.8, 0.8))
```

\raggedright
\scriptsize

Definition: `r russell_3000_data_def %>% filter(var_name == "roe") %>% select(var_def)`.
Data: Russel 3000 U.S. Index firms, fiscal years 2014-2016, as provided by the
ExPanDaR R package, https://joachim-gassen.github.io/ExPanDaR/.

## Data Visualization

Data visualization means mapping data on aesthetics
\vspace{24pt}
\begin{center}
\includegraphics[width=0.8\textwidth]{media/fdv_fig_2-1.png} 
\end{center}


## Data visualization for exploring data

* Adapt a structured workflow of analysis
* Visuals do not need to be nice but you should be able to produce them quick
* This is were interactive graphics really shine
* Shamelessly self-advertising example below

\vspace{24pt}

\footnotesize
```{r, eval = FALSE}
library(ExPanDaR)
ExPanD(worldbank, cs_id = "country", ts_id = "year", 
       var_def = worldbank_var_def)
```
\normalsize


## Data visualization for presenting data

* What is your story?
* Who is your target audience (academics in your field?, academics in other 
fields?, scientific media?, general audience?)
* What are the restrictions of your communication channel (paper?, slide 
presentation?, web site? social web?)
* Produce the best that you can do, factoring in the opportunities and
limitations generated by the points above 


## An example for data presentation

* We want to provide a tweet-able visual that communicates the sluggish development
of African economies in the last 50 years
* The main point is to create awareness that the income gap between African 
economies and North-american/European economies is not closing over time while
Asian economies have been catching up
* Of course we want our visual to be fully reproducible


## Zooming in on geographic regions

```{r, echo = FALSE, cache = TRUE, fig.width = 4, fig.height = 3, out.width="\\textwidth"}
eu15 <- c("Austria", "Belgium", "Denmark", "Finland", "France", "Germany", 
          "Greece", "Ireland", "Italy", "Luxembourg", "Netherlands", "Portugal", 
          "Spain", "Sweden", "United Kingdom")


east_asia <- c("Brunei Darussalam", "Cambodia", "China", "Indonesia", 
               "Korea, Rep.", "Lao PDR", "Malaysia", "Mongolia", "Myanmar", 
               "Papua New Guinea", "Philippines", "Singapore",  "Thailand", 
               "Timor-Leste", "Vietnam")


africa <- c("Algeria", "Djibouti", "Egypt, Arab Rep.", "Libya", "Morocco", 
            "Oman", "Tunisia", "Angola", "Benin", "Botswana", 
            "Burkina Faso", "Burundi", "Cabo Verde", "Cameroon", 
            "Central African Republic", "Chad", "Comoros", 
            "Congo, Dem. Rep.", "Congo, Rep.", "Cote d'Ivoire", 
            "Equatorial Guinea", "Eritrea", "Eswatini", "Ethiopia", 
            "Gabon", "Gambia, The", "Ghana", "Guinea", "Guinea-Bissau", 
            "Kenya", "Lesotho", "Liberia", "Madagascar", "Malawi", "Mali", 
            "Mauritania", "Mauritius", "Mozambique", "Namibia", 
            "Niger", "Nigeria", "Rwanda", "Sao Tome and Principe", 
            "Senegal", "Seychelles", "Sierra Leone", "South Africa", 
            "Sudan", "Tanzania", "Togo", "Uganda", "Zambia", "Zimbabwe")

north_america <- c("Canada", "United States")

wb <- wb %>%
  mutate(geo_region = case_when(
    country %in% eu15 ~ "EU 15",
    country %in% east_asia ~ "East Asia",
    country %in% africa ~ "Africa",
    country %in% north_america   ~ "North America",
    country == "Japan" ~ "Japan",
    TRUE ~ "Other"
  )) %>%
  select(1:3, geo_region, everything())

df <- wb %>%
  filter(geo_region != "Other") %>%
  mutate(year = as.numeric(as.character(year))) %>%
  group_by(geo_region, year) %>%
  summarise(pop_weighted_gdp_capita = weighted.mean(gdp_capita, population))
    
ggplot(df, aes(x = year, y = pop_weighted_gdp_capita, color = geo_region)) +
  geom_line() + theme_minimal()
```


## Base adjusting

```{r, echo = FALSE, cache = TRUE, fig.width = 4, fig.height = 3, out.width="\\textwidth"}
df <- df %>%
  arrange(geo_region, year) %>%
  group_by(geo_region) %>%
  mutate(base_line = first(pop_weighted_gdp_capita),
         rel_pop_weighted_gdp_capita = 100*pop_weighted_gdp_capita/base_line)
  
ggplot(df, aes(x = year, y = rel_pop_weighted_gdp_capita, color = geo_region)) +
  geom_line() + theme_minimal()
```


## Reducing information

```{r, echo = FALSE, cache = TRUE, fig.width = 4, fig.height = 3, out.width="\\textwidth"}
df <- wb %>%
  filter(geo_region != "Other") %>%
  mutate(geo_region = ifelse(geo_region %in% c("EU 15", "Japan", "North America"),
                             "Developed Economies (EU 15, North America, Japan)",
                             geo_region)) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  group_by(geo_region, year) %>%
  summarise(pop_weighted_gdp_capita = weighted.mean(gdp_capita, population)) %>%
  ungroup() %>% 
  pivot_wider(id_cols = year, names_from = geo_region, 
              values_from = pop_weighted_gdp_capita) %>%
  mutate(Africa = `Developed Economies (EU 15, North America, Japan)`/Africa,
         `East Asia` = `Developed Economies (EU 15, North America, Japan)`/`East Asia`) %>%
  select(year, Africa, `East Asia`) %>% 
  pivot_longer(-year, names_to = "geo_region", values_to = "income_multiple")

ggplot(df, aes(x = year, y = income_multiple, color = geo_region)) +
  geom_line() + theme_minimal()
```


## Catering to viewing habits (positive slopes --- good)

```{r, echo = FALSE, cache = TRUE, fig.width = 4, fig.height = 3, out.width="\\textwidth"}
df <- wb %>%
  filter(geo_region != "Other") %>%
  mutate(geo_region = ifelse(geo_region %in% c("EU 15", "Japan", "North America"),
                             "Developed Economies (EU 15, North America, Japan)",
                             geo_region)) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  group_by(geo_region, year) %>%
  summarise(pop_weighted_gdp_capita = weighted.mean(gdp_capita, population)) %>%
  ungroup() %>% 
  pivot_wider(id_cols = year, names_from = geo_region, 
              values_from = pop_weighted_gdp_capita) %>%
  mutate(Africa = Africa/`Developed Economies (EU 15, North America, Japan)`,
         `East Asia` = `East Asia`/`Developed Economies (EU 15, North America, Japan)`) %>%
  select(year, Africa, `East Asia`) %>% 
  pivot_longer(-year, names_to = "geo_region", values_to = "income_rel_to_dev_econ")

ggplot(df, aes(x = year, y = income_rel_to_dev_econ, color = geo_region)) +
  geom_line() + theme_minimal() 
```


## Nicing up

```{r, echo = FALSE, cache = TRUE, fig.width = 5, fig.height = 5, fig.align = "right", out.height="0.8\\textheight", fig.showtext = TRUE}
# This uses cobweb to freely place titles and captions
# See: https://stackoverflow.com/questions/25401111/left-adjust-title-in-ggplot2-or-absolute-position-for-ggtitle/47625672#47625672

p <- ggplot(df, aes(x = year, y = income_rel_to_dev_econ, color = rev(geo_region))) +
  geom_line() + 
  geom_line(size = 1.5) + 
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    breaks = c(0, 0.05, 0.10, 0.15), limits = c(0, NA), 
    minor_breaks = NULL,
    sec.axis = dup_axis(
      breaks = df$income_rel_to_dev_econ[df$year == 2017],
      labels = c("Africa", "East Asia"),
      name = NULL)
  ) +
  scale_color_manual(values = c(trr266_petrol, trr266_yellow)) + 
  xlab("Year") +
  ylab("Income relative to developed economies") +
  labs(
    x = "Year",
    y = "Income relative to developed economies",
    title = "\n",
    subtitle = "\n\n\n",
    caption = "\n\n\n\n\n"
  ) +
  theme(
    text = element_text(family = "PTSans"),
    panel.background = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey90"),
    axis.ticks = element_blank(),
    legend.position = "none",
    axis.text.y.right = element_text(margin = margin(0, 0, 0, 0)),
    axis.title = element_text(hjust = 1),
    plot.subtitle=element_text(color = "gray50")
  )

title_str <- "Africa is not catching up" 
subtitle_str = paste(
  "Average income per capita is deteriorating in Africa over time,",
  "relative\nto developed economies in Europe, Japan and North America",
  "but\nalso relative to economies in east Asia."
)
caption_str = paste(
  "Income relative to developed economies is the population weighted",
  "mean of\ncountry-level GDP measures from the respective region",
  "relative to countries\nfrom developed economies (Canada, EU 15,",
  "Japan, U.S.).\nData: World Bank.",
  "Code: https://github.com/joachim-gassen/sposm."
)
ggdraw(p) + 
  draw_text(title_str, x = 0.01, y = 0.95, hjust = 0, vjust = 1, 
            family = "PTSans") +
  draw_text(subtitle_str, x = 0.01, y = 0.90, hjust = 0, vjust = 1, 
            color = "grey50", size = 11, lineheight = 1, family = "PTSans") + 
  draw_text(caption_str, x = 0.01, y = 0.15, hjust = 0, vjust = 1, 
            color = "grey80", size = 10, lineheight = 1, family = "PTSans")
```
