---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Statistical programming languages: An overview
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```

## Time table October 10

``` {r ttable, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-10-10")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>%
  row_spec(6, background = trr266_lightpetrol)
```

## Obvious aspects to consider when picking the 'right' SPL

* User base
* Range of features
* Ease of learning
* Performance
* Extendability
* License model and pricing


## Not so obvious aspects

* Portability across platforms
* Portability across time
* Verifiability of algorithms
* Quality of documentation
* Interoperability with other languages
* Interoperability with RDBMS
* Dynamic output creation (HTML/Javascript)
* User community


## My take on the Top 4: Julia, Python, R and Stata 

* Julia: 
  - Pros: Fast, open source, expert user base
  - Cons: Small user base, few packages
* Python:
  - Pros: General purpose, open source, relatively easy to learn, many machine learning packages, large user base
  - Cons: Packaging system, statistic packages have limited interoperabilty, object orientation feels alien when working with data
* R:
  - Pros: Focused on data science, open source, packaging system, interoperability, graphics system
  - Cons: Not really easy to learn, tidyverse helps though
* Stata
  + Pros: Easy to learn, very broad user base in economics, most statisitical methods are quickly implemented
  - Cons: Commercially licensed and closed source, inflexible programming environment


## Activity: Compare our code solution

* Let's compare our solutions
* Do our samples differ? If yes: why? 
* Whose code is the fastest? 
* Whose code is the most readable?
