---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Tidy data scraping
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
library(tidyverse)

opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```


## Time table Monday, February 17

``` {r ttableMon, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-02-17")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>%
  row_spec(1:2, background = trr266_lightpetrol)
```


## Time table Tuesday, February 18

``` {r ttableTue, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-02-18")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) 
```


## Parsing an HTML table I

\begin{center}
\includegraphics[width=\textwidth]{media/scraping_01.png} 
\end{center}


## Parsing an HTML table II

\begin{center}
\includegraphics[width=\textwidth]{media/scraping_02.png} 
\end{center}


## Parsing an HTML table III

\footnotesize
```{r parseTable, eval=FALSE}
library(tidyverse)
library(rvest)

url_sp500_const <- paste0(
  "https://en.wikipedia.org/wiki/",
  "List_of_S%26P_500_companies"
)

url_sp500_const %>%
  read_html() %>%
  html_node(xpath = '//*[@id="constituents"]') %>%
  html_table() -> sp500_constituents_raw
```
\normalsize


## Retrieving local URLs from within table I

\begin{center}
\includegraphics[width=\textwidth]{media/scraping_03.png} 
\end{center}


## Retrieving local URLs from within table II

```{r parseLinks, eval=FALSE}
url_sp500_const %>%
  read_html() %>%
  html_node(xpath = '//*[@id="constituents"]') %>%
  html_nodes("td:nth-child(2) a") %>% 
  html_attr("href")-> links
```


## Scraping ill-structured tables I

\begin{center}
\includegraphics[width=\textwidth]{media/scraping_04.png} 
\end{center}


## Scraping ill-structured tables II

\footnotesize
```{r infoBox, eval=FALSE}
xml_data <- read_html(url) %>%
  html_node('#mw-content-text div table.infobox.vcard')

xml_data  %>%
  html_table(fill = TRUE) %>%
  rename(tag = X1, content = X2) %>%
  filter(tag != "")
```
\normalsize


## Another approach to scraping

\begin{center}
\includegraphics[height=0.8\textheight]{media/scraping_05.png} 
\end{center}


## This won't work

\footnotesize
```{r scapingFail, error=TRUE}
library(rvest)

url_bt_open_data <- "https://www.bundestag.de/services/opendata"
url_bt_open_data %>%
  read_html() %>%
  html_node(
    xpath = '//*[@id="bt-collapse-543410"]/div[1]/div/div/div[1]/table'
  ) %>% html_table() -> pp_table
```
\normalsize

Reason: The web page is being created dynamically by JavaScript (or similar) 


## The idea of headless browsing: meet Selenium

\begin{center}
\includegraphics[width=0.1\textwidth]{media/selenium_logo.png} 
\end{center}
\vspace{24pt}
- Selenium offers a way to script a web browser so that data can be scraped
using the navigation that a web page provides

- Allows for various web browser and all sorts of user web browser interaction

- Uses a docker container holding the actual browser environment

- See `code/btag_open_data_scrape_data.R` for a demonstration on how to use it


## Parsing XML Data

\begin{center}
\includegraphics[height=0.7\textheight]{media/xml_tree_navigate.pdf} 
\end{center}
\footnotesize
Gaston Sanchez, https://github.com/gastonstat/tutorial-R-web-data
\linebreak
CC BY-NC-SA 4.0
\normalsize


## My task ...

- Develop a function that parses the XML files of the Plenarprotokolle of the
19th Wahlperiode to extract all speaches into a tidy data structure

- Implement some basic test routines verifying that the code works

- See `code/btag_open_data_scrape_data.R` and 
`code/test/test_btag_open_data_scrape_data.R`