---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Debugging tools
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
library(ggmap)
opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```

## Time table October 11

``` {r ttable, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-10-11")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>%
  row_spec(3, background = trr266_lightpetrol)
```


## Disclaimer

\center Some of the following is borrowed from chapter 22 of 
Hadley Wickham (2019): Advanced R,
https://adv-r.hadley.nz


## Principle

Finding your bug is a process of confirming the many things that you believe are true — until you find one which is not true.

`r quote_footer("—-- Norm Matloff")`

\vspace{24pt}

* Google the error message
* Make the bug repeatable by creating a reproducible example
* Figure out where it is
* Fix it and test it


## Most bugs are simple

* Typos are paramount and sometimes hard to spot (a good editor with syntax 
high-lightening helps!)
* The best way to avoid bugs is do adopt a readable coding style
* Debugging and unit testing go hand in hand (we will talk about unit testing
in the February block)
* Do not ignore messages and warnings that your code throws at you. They are 
there for a reason,


## Error messages are not always self-explanatory but almost always informative

\footnotesize
```{r, error = TRUE, collapse = TRUE}
count_distinct_obs <- function(df, ...) {
  as_tibble(df) %>%
    select(...) %>%
    distinct() %>%
    nrow()
}

read_csv("../data/sub.csv") %>%
  count_distinct_obs[cik, name] -> count_sec_reg
```
\normalsize


## Debugging tools

- Traceback
- Debug on error
- Breakpoints
- Logging
- Dump analysis

\vspace{12pt}
\center

Let's do a case: Take a look at `code/calc_state_of_incorp_distance_bugged.R`

## Corner cases

\scriptsize
```{r, collapse = TRUE}
# This function returns a p % sample of a vector v. It always 
# chooses at least q observations to avoid small samples

sample_my_sample <- function(v, p, q) {
  n <- ceiling(max(p * length(v), q))
  do_replace <- (n > length(v))
  sample(v, n, do_replace)
}

for (i in 1:10) {
  print(sample_my_sample(i:10, 0.5, 3))
}
```
\normalsize


## Code meets data

\footnotesize
```{r, collapse = TRUE, fig.align="center", fig.height = 3, fig.width = 4, out.height = "0.5\\textwidth"}
read_csv("../raw_data/stock_price.csv") %>% 
  ggplot(aes(x = date, y = adj_prc)) + 
  geom_line() + theme_minimal()
```
\normalsize


## Calculate weekly mean return and weekly return variance

\footnotesize
```{r, collapse = TRUE}
read_csv("../raw_data/stock_price.csv") %>%
  mutate(return = (adj_prc - lag(adj_prc))/lag(adj_prc)) %>%
  summarise(mn_return = mean(return, na.rm = TRUE),
            var_return = sd(return, na.rm = TRUE)^2)

## Correct values
## # A tibble: 1 x 2
##   mn_return var_return
##       <dbl>      <dbl>
## 1  0.000434  0.0000247
```
\normalsize