---
title: |
  | Statistical Programming 
  | and Open Science Methods
subtitle: |
  | Relational databases and the concept of normalized data
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
date:  |
  | `r loc <- Sys.getlocale(category = "LC_TIME"); Sys.setlocale("LC_TIME", "C"); fdate <- format(Sys.time(), '%B %d, %Y'); Sys.setlocale("LC_TIME", loc); fdate`
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
#devtools::install_github("bergant/datamodelr")
library(datamodelr)
opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```

## Time table October 11

``` {r ttable, echo = FALSE}
df <- sposm19_time_table %>%
  filter(day(dtime) == day("2019-10-11")) %>%
  mutate(dtime = format(dtime, "%H:%M"))

breaks <- which(str_detect(df$title, "Lunch") | str_detect(df$title, "Coffee"))

names(df) <- c("When?", "What?")
kable(df, booktabs = TRUE, linesep = "")  %>%
  kable_styling(latex_options = c("striped", "scale_down"),
                stripe_index = breaks) %>%
  row_spec(5, background = trr266_lightpetrol)
```


## Disclaimer

\center Some of the following --- in particular the nice-looking figures ---
are borrowed from Garrett Grolemund and Hadley Wickham (2017):
R for Data Science, https://r4ds.had.co.nz


## Think about the following scenarios

- You have a project that collects data via a web form/survey
- You have a time-intensive process that you want to multi-thread across several
platforms
- You have big data that you need to query but the amount of data that you
actually need for your analysis is relatively small
- You are reusing data for multiple projects

\vspace{12pt}

In all these cases you might need a relational database management 
system (RDBMS)


## Our SEC data has a relational structure \dots

```{r sec-datamodel, cache = TRUE, echo = FALSE, fig.align="center", out.height="0.8\\textheight"}
for(df_name in c("sub", "tag", "num", "pre")) {
  assign(df_name, read_csv(paste0("../data/", df_name, ".csv")))
}
sub <- sub %>% select(adsh, cik, name, sic, countryba, form, period, fp)
sec_dm <- dm_from_data_frames(sub, tag, num , pre)

sec_dm <- dm_add_references(
  sec_dm,
  num$adsh == sub$adsh,
  num$tag == tag$tag,
  num$version == tag$version,
  pre$adsh == sub$adsh,
  pre$tag == tag$tag,
  pre$version == tag$version,
  pre$adsh == num$adsh,
  pre$tag == num$tag,
  pre$version == num$version
)
graph <- dm_create_graph(sec_dm, rankdir = "LR", col_attr = c("column", "type"))
dm_render_graph(graph)
```


## Understanding the concept of relations: Fundamentals

* Observations are stored in rows, variables are stored in columns

* Only one observation per row, only one variable per column

* Each observation has a unique primary key (might be a set of keys)


\begin{center}
\includegraphics[width=\textwidth]{media/r4ds_fig_12-1.png} 
\end{center}



## Primary and foreign keys

* A primary key (or set of keys) identifies an observation in a given dataset/table

* A foreign key (or a set of keys) identifies an observation in _another_ dataset/table

* This implies that a foreign key is not necessary unique (while the primary key is)

* In is a very useful habit to organize joins so that you use the foreign key of the first (left) and the primary key of the second (right) dataset


## Joining datasets: 1:1 joins

* There are various type of joins: The inner join

\vspace{36pt}

\begin{center}
\includegraphics[width=0.5\textwidth]{media/r4ds_join-inner.png} 
\end{center}


## Joining datasets: 1:1 joins

* The outer joins

\begin{center}
\includegraphics[width=0.4\textwidth]{media/r4ds_join-outer.png} 
\end{center}


## Joining datasets: 1:1 joins

* The 1:1 joins combined
 
\vspace{36pt}
 
\begin{center}
\includegraphics[width=\textwidth]{media/r4ds_join-venn.png} 
\end{center}


## Joining datasets: 1:n joins

* Joins one observation from one dataset to potentially many observations from another dataset. Can and in most cases will enlarge the sample size

\vspace{36pt}
 
\begin{center}
\includegraphics[width=0.5\textwidth]{media/r4ds_join-one-to-many.png} 
\end{center}


## Joining datasets: n:m joins (the wrong way)

* Directly joining two datasets that form a n:m relation is not feasible as observations are no longer uniquely defined in _both_ tables.

\vspace{36pt}
 
\begin{center}
\includegraphics[width=0.5\textwidth]{media/r4ds_join-many-to-many.png} 
\end{center}


## Joining done right: A "case study" about normalization

Assume that we run an event for the TRR 266. People register for the
event. To keep track of registration, we set up the following data
set.

\vspace{24pt}

\footnotesize
```{r}
participant = tibble(
  name = c("Andreas", "Tina", "Xhi"),
  email = c("andi@gmail.com", "tina@hu-berlin.de", "xhi@upb.de"),
  project = c("A02", "B02", "A05, B08")
)
```
\normalsize

What do you think about this approach?


## Oh, OK. You want a key

All good now?

\vspace{24pt}

\footnotesize
```{r}
participant <- tibble(
  participant_id = 1:3,
  name = c("Andreas", "Tina", "Xhi"),
  email = c("andi@gmail.com", "tina@hu-berlin.de", "xhi@upb.de"),
  project = c("A02", "B02", "A05, B08")
)
```
\normalsize


## `project` is not atomic

`participants$project[3] ("A05, B08")` contains two observations. How to fix this? Maybe by:

\vspace{24pt}

\footnotesize
```{r}
participant <- tibble(
  participant_id = 1:3,
  name = c("Andreas", "Tina", "Xhi"),
  email = c("andi@gmail.com", "tina@hu-berlin.de", "xhi@upb.de"),
  project1 = c("A02", "B02", "A05"),
  project2 = c(NA, NA, "B08")
)
```
\normalsize

Now everything is atomic (One value per cell). Technically, we have reached the first normal form of a relational database. Is it a smart solution?


## Addressing insertion anomalies

No. We created an _insertion anomaly_. If somebody happens to be on 
three projects, we would be unable to enter this information in our
data model. Let's address this.

\vspace{12pt}

\footnotesize
```{r}
participant <- tibble(
  participant_id = 1:3,
  name = c("Andi", "Tina", "Xhi"),
  email = c("andi@gmail.com", "tina@hu-berlin.de", "xhi@upb.de")
)

people_project <- tibble(
  participant_id = c(1, 2, 3, 3),
  project_id = c("A02", "B02", "A05", "B08")
)
```
\normalsize

## Adding data

We are happy about our new data model but we do not know what these
fancy project IDs stand for. Why not add project names?
\vspace{12pt}

\footnotesize
```{r}
people_project$project_title <- c(
  "Transparency Effects of Organizational Innovations",
  "Private Firm Transparency",
  "Accounting for Tax Complexity",
  "Tax Burden Transparency"
) 
```
\normalsize

We are happy. Right?


## Meet the second normal form (2NF)

We just introduced data into your model that only depends on `project_id`, not on `participant_id`. As both together define the key of the dataset ``people_in_projects`, we violated the requirements of second form normality: All non key data in a table have to depend on _all_ keys, not on a subset.

This is important as a violation creates an _update anomaly_. 
In our case: When we change a project title, we potentially 
need to change in it multiple instances in your data.

Let's fix this.

\footnotesize
```{r}
project <- tibble(
  project_id = people_project$project_id,
  project_title = people_project$project_title
)

people_project <- people_project %>%
  select(- project_title)
```
\normalsize


## Deleting data \dots

We are very proud of ourselves. Now Tina calls. She can't make it to
our event. No problem. We simply delete her entry and send her an email
to acknowledge that she is no longer registered. 

\vspace{24pt}

\footnotesize
```{r}
participant <- participant %>%
  filter(participant_id != 2)
```
\normalsize

But, hey, where did her email go? 


## The third normal form

Our `participant` table is subject to a _deletion anomaly_ as it is not
in the third normal form (3NF). When we delete data, we also delete
data that is still informative (the email in our case).

The third normal form requires on top of the second normal form that
there are no _transitional dependencies_ in our data, meaning that all
the non-keys are directly depending on the keys and not on some other 
variable that in turn depends on the keys.

## Time for a last change

\footnotesize
```{r}
people <- tibble(
  people_id = 1:3,
  name = c("Andi", "Tina", "Xhi"),
  email = c("andi@gmail.com", "tina@hu-berlin.de", "xhi@upb.de")    
)

people_event <- tibble(
  event_id = c(1, 1),
  people_id = c(1, 3)
)

event <- tibble(
  event_id = 1,
  name = "Our super nice event"
)

people_project <- people_project %>%
  rename(people_id = participant_id)

rm(participant)
```
\normalsize


## Our data model and voila: two n:m relations

```{r {case-dm}, echo = FALSE, fig.height = 3, fig.width = 4, fig.align = "center", out.width="0.6\\textwidth", cache = TRUE}
my_dm <- dm_from_data_frames(
  people_project, project, people, people_event, event
)
my_dm <- dm_auto_create_keys_and_references(my_dm)
graph <- dm_create_graph(my_dm, rankdir = "TB")
dm_render_graph(graph)
```


## Using external RDBMS for storage

* Using database solutions external to your own programming code has many advantages
* Your can access your data from various platforms
* RDBMS take care of potential race conditions when concurrent access is feasible
* Data integrity is maintained across applications
* Query speed (at least in many cases)


## A simple example: SQLite

* `SQLite` is a light-weight single-user file-based RDBMS
* It qualifies as ACID: Transactions are _atomic_, _consistent_, _isolated_, and _durable_. 
* Using such a RDBMS does not avoid all race conditions (concurrent data changes leaving data in inconsistent state) but significantly reduces the situations where race conditions can incur

\vspace{12pt}

Larger scale RDBMS that can be outsourced to different servers include
_PostgreSQL_, _MySQL_, and _MariaDB_

## Moving our SEC data to SQLite

\footnotesize
```{r, cache = TRUE, message = FALSE, collapse = TRUE}
library(DBI)
if (file.exists("../data/sec.sqlite3")) 
  unlink("../data/sec.sqlite3")
con <- dbConnect(RSQLite::SQLite(), "../data/sec.sqlite3")
for(df_name in c("sub", "tag", "num", "pre")) {
  dbWriteTable(
    con, df_name, 
    read_csv(paste0("../data/", df_name, ".csv"))
  )
}
dbListTables(con)
dbDisconnect(con)
```
\normalsize


## It pays performance-wise (100 runs each)

```{r, eval = FALSE, echo = FALSE}
# This code is non-portable as I hard-coded the links to Python and Stata.
# Sorry for being lazy. And as it runs several hours (thank you, Stata),
# I stored the data permanently.

library(microbenchmark)

mbm_100 <- microbenchmark(
  r_flat_file = source("code/extract_revenues_sec_fin_stat_data.R"),
  python_flat_file = system("/Users/joachim/anaconda3/bin/python3 code/extract_revenues_sec_fin_stat_data.py"),
  stata_flat_file = system("/Applications/Stata/StataSE.app/Contents/MacOS/StataSE -e code/extract_revenues_sec_fin_stat_data.do"),
  r_sqlite = source("code/extract_revenues_sec_fin_stat_data_db.R"),
  times = 100
)
saveRDS(mbm_100, "raw_data/microbenchmark_extract_revenues_sec_fsd.RDS")

```

```{r, echo = FALSE, fig.width=4, fig.height=3, out.width="0.9\\textwidth"}
mbm_100 <- readRDS("../raw_data/microbenchmark_extract_revenues_sec_fsd.RDS") %>%
  mutate(seconds = time/1e9) 

levels(mbm_100$expr) <- c("R\n(CSV data)", "Python 3\n(CSV data)", 
                          "Stata\n(CSV Data)", "R\n(SQLite)") 

ggplot(data = mbm_100 %>% filter(expr != "Stata\n(CSV Data)"), 
       aes(x = expr, y = seconds, color = expr)) +
  geom_boxplot(outlier.shape = NA, fill = NA) +
  geom_point(aes(fill = expr), alpha = 0.5, size = 1,
             position=position_jitter(width = 0.2)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Approach", y = "Execution time in seconds")
```

Note: Runs for Stata (CSV data) executed from plot (mean = 
`r format(mean(mbm_100$seconds[mbm_100$expr == "Stata\n(CSV Data)"]), digits = 0)`)
